---
title: "Functional heterogeneity in Mesenchymal stromal cell (MSC) subtypes."
author: "Emma Rand"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
  bookdown::word_document2: default
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.retina = 3)
```

```{r pkg}
library(tidyverse)
library(plotly)
library(heatmaply)
library(GGally)
```

```{r functions}
# data_summary calculates summary statistics (mean, sample size, standard deviation, standard error) for a numeric variable. Used to summarise the protein abundances for each cell line
source("functions/data_summary.R")
```

```{r labels}
lines = c("Y1015", 
          "Y101",
          "Y201",
          "Y202",
          "Y102")
functionality = c("Super Osteogenic",
                  "Osteogenic",
                  "Multipotent",
                  "Immunomodulatory",
                  "Immunomodulatory")

```

```{r pallete}
# from viridis::inferno
# "Y1015" dark purple
# "Y101" mid purple
# "Y201" dark pink
# "Y202" orange
# "Y102" yellow
line_colours <- viridis::inferno(7, 
                                 begin = 0.15,
                                 end = 0.85)[c(1, 2, 4, 6, 7)]
```

```{r banner-data}
df <- data.frame(lines = factor(lines, levels = lines), 
                 y = 1, 
                 functionality = functionality)
```

```{r banner, fig.height=.5, out.width = "100%", fig.align='center'}

df %>% 
  ggplot(aes(lines, y, fill = lines)) +
  geom_col(width = 1) +
  geom_label(aes(x = lines, y = y*2/3, label = lines),
             fill = "white",
             size = 2,
             label.size = 0,
             label.padding = unit(0.2, "lines")) +
  geom_label(aes(x = lines, y = y*1/3, label = functionality),
             fill = "white",
             size = 2,
             label.size = 0,
             label.padding = unit(0.2, "lines")) +
  scale_fill_manual(values = line_colours) +
  theme_void() +
  theme(legend.position = "none")
  

```

# Introduction {#introduction}

Mesenchymal stromal cells (MSCs) are a subset non-hematopoietic adult stem cells that originate from the mesoderm. They have the capacity for self-renewal and differentiation into osteoblasts, adipocytes and chondrocytes making them important for the development of cell-based therapies in regenerative medicine. However, populations of MSCs are heterogeneous with respect to their differentiation capacity [@Costa2020-ze; @Phinney2012-yw] and some exhibit immundomodulatory properties unhelpful for cell-based therapies. Understanding subtype heterogeneity is key for the development of efficacious therapies. [The Genever Group](https://www.geneverlab.info/) has developed a model of this heterogeneity using has a number of immortalised clonal MSC lines which comprises five subtypes [@Stone2019-wr; @Kay2019-ec]. The phenotypes of these subtypes is as follows:

-   Y201 : the MSC stereotype which differentiate into osteoblasts, adipocytes and chondrocytes
-   Y101 : osteogenic capacity
-   Y202 : immundomodulatory capacity
-   Y102 : immundomodulatory capacity
-   Y101.5 : enhanced osteogenic capacity

The data are mass spectrometry data of the soluble protein fraction from five immortalised mesenchymal stromal cell (MSC) lines.

# Methods {#methods}

## Data description {#methods-data-description}

There are two data files: [Y101_Y102_Y201_Y202_Y101-5.csv](data-raw/Y101_Y102_Y201_Y202_Y101-5.csv) and [comparison_p\_and_q.csv](data-raw/comparison_p_and_q.csv)

The data in Y101_Y102_Y201_Y202_Y101-5.csv are normalised protein abundances. Each row is a protein and there are three replicates for each subtype which are arranged in columns. Also in the file are columns for:

-   the protein accessions: the top hit is first, Human sequences are prefixed with `1::`, bovine with `2::`
-   the number of peptides used to identify the protein
-   the number of unique peptides used to identify the protein
-   a measure of confidence in that identification
-   the maximum fold change between the mean abundances of two cell lines (i.e., highest mean / lowest mean)
-   a $p$ value for a comparison test between the highest mean and lowest mean
-   $q$ the false discovery rate, a $p$ value corrected by the Benjamini--Hochberg procedure [@Benjamini1995-yj].
-   a measure of the power of that test
-   the cell line with the highest mean
-   the cell line with the lowest mean
-   the protein mass
-   a description of the protein
-   a binary indicator variable for whether at least two peptides were detected for a protein.

The data in comparison_p\_and_q.csv give the p and q values for pairwise comparisons between cell line.

```{r data-import}
# file
filesol <- "data-raw/Y101_Y102_Y201_Y202_Y101-5.csv"
# Column names are spread over three rows but are primarily in the third row.
# skip first two lines
sol <- read_csv(filesol, skip = 2) %>% 
  janitor::clean_names()

file <- "data-raw/comparison_p_and_q.csv"
p_q <- read_csv(file) %>% 
  janitor::clean_names()
```

## Data preparation {#methods-data-preparation}

```{r data-filter}
# Remove irrelevant rows

# bovine serum proteins removed - these are from medium on which the cells were grown 
# proteins for which fewer than 2 peptides were detected
sol <- sol %>% 
  filter(str_detect(description, "OS=Homo sapiens")) %>% 
  filter(x1pep == "x")
num_prot <- length(sol$accession) 

p_q <- p_q %>% 
  filter(str_detect(description, "OS=Homo sapiens")) %>% 
  filter(x1pep == "x")

```

```{r add-useful-columns}
# Extract important information 'hidden' in other columns to columns of their own

# extract the genename from the description and put it in a column.
# genename is after GN= in the description until a white space
sol <- sol %>%
  mutate(genename =  str_extract(description,"GN=[^\\s]+") %>% 
           str_replace("GN=", ""))

p_q <- p_q %>%
  mutate(genename =  str_extract(description,"GN=[^\\s]+") %>% 
           str_replace("GN=", ""))


# Add a column, protid, for the top protein identifier. 
# This is the first Uniprot ID after the "1::" in the accession column.
sol <- sol %>%
  mutate(protid =  str_extract(accession, "1::[^;]+") %>% 
           str_replace("1::", ""))

```

```{r create-long-sol2}
# Create a second dataframe, sol2 in which the protein abundances are in a single column
# these are the columns beginning with a y
# the column names which indicate the cell lineage and replicate
# good to a column named lineage_rep which is then split into separate columns
# All the other variables are retained.
sol2 <- sol %>% pivot_longer(names_to = "lineage_rep",
                             values_to = "abundance",
                             cols = starts_with("y")) %>%
  extract(lineage_rep,
          c("line", "rep"),
          "(y[0-9]{3,4})\\_([a-c])")
```

```{r summary-sol}
# add a dataframe for the mean, se, sd, and n abundance for each cell line protein combination
sol_summary <- data_summary(sol2, 
                            measure = abundance, 
                            group1 = line,
                            group2 = protid)
# add a dataframe for the mean, se, sd, and n abundance for each cell line 
sol_summary_line <- data_summary(sol2, 
                            measure = abundance, 
                            group1 = line)
sol_summary_line$line <- factor(toupper(sol_summary_line$line), levels = lines)
```

```{r augment-sol}
# add the protein-line summary information to the main dataset
# requires protein-line summary info to be spread
sol_summary <- sol_summary %>% 
  pivot_wider(names_from = line,
              values_from = c(mean, n, sd, se))
# merge the dataframes
sol <- merge(sol, sol_summary, by = "protid")
```

```{r}
# calculate the log fold changes for each of the comparisons
# we have p values for. This is all the pairwaise comparisions
# except that y1015 is commpared only to y101, no others
sol <- sol %>% 
  mutate(fc_y101_y1015 = log2(mean_y101/mean_y1015), # have p
         fc_y101_y102 = log2(mean_y101/mean_y102), # have p
         fc_y101_y201 = log2(mean_y101/mean_y201), # have p
         fc_y101_y202 = log2(mean_y101/mean_y202), # have p
         ##
         fc_y102_y201 = log2(mean_y102/mean_y201), # have p
         fc_y102_y202 = log2(mean_y102/mean_y202), # have p
         ##
         fc_y201_y202 = log2(mean_y201/mean_y202)) # have p
```

I used R [@R-core] with **`tidyverse`** packages [@R-tidyverse] to import and process the raw data. Bovine serum proteins from the medium on which the cells were grown and those for which fewer than two peptides were detected were filtered out. Abundances were summarised for each line and for each line-protein combination. The log fold change between cell lines for each protein was calculated. See Equation \@ref(eq:logfc). 


\begin{equation}
FC_{jj'} = log_{2}\left(\frac{mean_{j}}{mean_{j'}}\right)
(\#eq:logfc)
\end{equation}

where: 

* $mean_{j}$ is the mean protein abundance for a given protein in cell line $j$  
* $mean_{j'}$ is the mean protein abundance for a given protein in another cell line $j'$  
* such that $FC_{jj'}$ is the $log_{2}$ fold change between line $j$ and line $j'$ for a given gene.


## Visualisation with Principal Components Analysis {#methods-visualisation}

```{r pca-preparation}
# PCA requires the proteins to be in columns and the samples to be in rows
# Transpose the columns containing the abundance values for each sample
tsol <- sol %>% select(starts_with("y")) %>%
  t() %>% 
  data.frame()

# Use the genenames in `sol` to name the columns in `tsol`
# I'm using the genenames rather than the protein ids because they
# are more recognisable
names(tsol) <- sol$genename 

# The sample names of sol have become the row names of tsol
# put these in a column 'sample
tsol$sample <- row.names(tsol)

# process the sample name so we have the cell lineage in one column and the replicate in another
tsol <- tsol %>% 
  extract(sample, 
          c("lineage","rep"),
          "(y[0-9]{3,4})\\_([a-c])")
```

```{r pca-run}
pca <- tsol %>% 
  select(-lineage, -rep) %>%
  prcomp(scale. = TRUE)
```

Principal Components Analysis was used for dimensionality-reduction to allow visualisation the distance between samples. It was conducted on protein abundances scaled to unit variance to prevent highly abundant proteins dominating the analysis. The package **`GGally`** [@GGally] was used to produced pairwise scatter plots of the first six principal components.

## Clustering and heatmaps {#methods-clustering-and-heatmaps}

```{r cluster-filter}
# select genes where at least one comparison differs significantly at the q < 0.01 level
row.names(sol) <- sol$genename
sol_sigq_0.01 <- sol %>% 
  filter(q_value <= 0.001) %>% 
  select(starts_with("y")) 
```

```{r cluster-prep}
# Matrix format
mat <- sol_sigq_0.01 %>% as.matrix()
```

Hierarchical clustering was conducted and heatmaps produced with the package **`heatmaply`** [@heatmaply] with interactivity provided by **`plotly`** [@plotly]. Hierarchical clustering used the complete linkage method and was performed for both proteins and samples on the proteins that differed significantly (at the $q < 0.01$ level) between at least two lines. The abundances were "$z$-score normalised" for each protein before clustering. $Z$-score normalisation is way to make comparisons between, and visualisations of, abundances on different scales. See Equation \@ref(eq:zscore).


\begin{equation}
z_{ijk} = \frac{x_{ijk} - \bar{x_{ij}}}{sd_{ij}}  
(\#eq:zscore)
\end{equation}

where:  

* $x_{ijk}$ is the abundance for protein $i$, cell line $j$ and replicate $k$  
* $\bar{x_{ij}}$ is the mean abundance for protein $i$ and cell line $j$ across replicates  
* $sd_{ij}$ is the standard deviation on the abundance for protein $i$ and cell line $j$ across replicates  
* such that $z_{ijk}$ is the $z$-score normalised abundance for protein $i$, cell line $j$ and replicate $k$

## Volcano plots {#methods-volcano-plots}

```{r volcano-prep}
sol <- merge(sol, p_q, by = "genename")
```

# Results {#results}

There were `r num_prot` human protein identified from more than two peptides and the total soluble protein content of each cell line was broadly similar. See figure \@ref(fig:line-abund-fig).

```{r line-abund-fig, fig.height=4, fig.width=4, fig.cap="(ref:line-abund-fig)"}
# ..
sol_summary_line %>% 
  ggplot() +
  geom_col(aes(x = line, y = mean, fill = line)) +
  geom_errorbar(aes(x = line,
                    ymin = mean - se,
                    ymax = mean + se),
                width = .5) +
  scale_x_discrete(labels = lines,
                   name = "") +
  scale_fill_manual(values = line_colours) +
  scale_y_continuous(name = "Protein Abundance (units)",
                     expand = c(0, 0)) +
  theme_classic() +
  theme(legend.position = "none")

```

<!-- figure legend -->

(ref:line-abund-fig) Mean soluble protein content for each cell line. Error bars are $\bar{x} \pm 1 s.e.$

<!-- end of fig legend -->


```{r}
venndf <- data.frame(gene = sol$genename,
                     Y101 = sol$mean_y101 > 0,
                     Y1015 = sol$mean_y1015 > 0,
                     Y102 = sol$mean_y102  > 0,
                     Y201 = sol$mean_y201 > 0,
                       Y202 = sol$mean_y202 > 0)

counts <- limma::vennCounts(venndf[-1])

limma::vennDiagram(counts,
                   circle.col = line_colours[c(1,5,3,4,2)],
                   lwd = 2,
                   mar = rep(0.5,4))
```



## Principal Components analysis {#results-principal-components-analysis}

```{r pca-report}
res <- summary(pca)[["importance"]]

percent1 <- res["Cumulative Proportion","PC1"] * 100
percent6 <- res["Cumulative Proportion","PC6"] * 100
```

```{r pca-vis}
pca_labelled <- data.frame(pca$x, 
                           lineage = factor(toupper(tsol$lineage),
                                            levels = lines))
# a then to do a scatterplot
pca_labelled %>% 
  select(lineage,
         PC1:PC6) %>%
  ggpairs(aes(color = lineage, fill = lineage),
          upper = NULL,
          columns = 2:7) + 
  scale_colour_manual(values = line_colours) +
  scale_fill_manual(values = line_colours) +
  theme_minimal()
  
```

## Clustering and heatmaps {#results-cluster-heat}

```{r}
n_gen_clusters <- 3
n_cell_clusters <- 5
p <- heatmaply(mat, 
               scale = "row",
               grid_color = "white",
               hide_colorbar = TRUE,
               k_col = n_cell_clusters,
               k_row = n_gen_clusters,
               label_names = c("Protein", "Sample", "Normalised Abundance"),
               fontsize_row = 7, fontsize_col = 10,
               labCol = toupper(colnames(mat)),
               labRow = rownames(mat),
               heatmap_layers = theme(axis.line = element_blank()))
```

## Volcano plots {#results-volcano-plots}

```{r}
sol %>% 
  ggplot(aes(x = fc_y101_y1015, 
             y = -log(q_y101_y1015),
             colour = abs(fc_y101_y1015) > 2 & q_y101_y1015 < 0.05)) +
  geom_point() +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
file <- "sessioninfo.md"
writeLines(capture.output(sessionInfo()), file)
```

## Word count {#word-count}

Word count calculated with wordcountaddin [@wordcountaddin]. The session information was written into a separate file rather than to the README this has been added to the wordcount.

```{r wordcount}
wc_main <- wordcountaddin:::text_stats("report/main.Rmd")[3] %>%
  str_extract("[0-9]+") %>% 
  as.numeric()
wc_readme <- wordcountaddin:::text_stats("README.md")[3] %>%
  str_extract("[0-9]+") %>%
  as.numeric()
wc_session <- wordcountaddin:::text_stats("sessioninfo.md")[3] %>%
  str_extract("[0-9]+") %>%
  as.numeric()

```

This document: `r wc_main`\
README: `r wc_readme`\
Session info: `r wc_session`\
**Total: `r wc_main + wc_readme + wc_session`**

# References {#references}
