---
title: "Functional heterogeneity in Mesenchymal stromal cell (MSC) subtypes."
author: "Emma Rand"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
  bookdown::word_document2: default
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.retina = 3)
```

```{r pkg}
library(tidyverse)      # import, tidying, summary, plotting
#library(plotly)        # import, tidying, summary, plotting
library(heatmaply)      # interactive heatmap
library(GGally)         # PCA visualisation
library(VennDiagram)    # protein set venn
library(patchwork)      # arranging ggplots
```

```{r functions}
# data_summary calculates summary statistics (mean, sample size, standard deviation, standard error) for a numeric variable. Used to summarise the protein abundances for each cell line
source("../functions/data_summary.R")
```

```{r labels}
# cell line name for figures and ordering of factor levels
lineage = c("Y1015", 
          "Y101",
          "Y201",
          "Y202",
          "Y102")
# broad functionality of each cell line, order to match 'lineage'
functionality = c("Super Osteogenic",
                  "Osteogenic",
                  "Multipotent",
                  "Immunomodulatory",
                  "Immunomodulatory")

```

```{r pallete}
# palette for cell lineage from viridis::inferno
# order to match 'lineage' and colour proximity reflects functionality proximity
# "Y1015" dark purple
# "Y101" mid purple
# "Y201" dark pink
# "Y202" orange
# "Y102" yellow
line_colours <- viridis::inferno(7, 
                                 begin = 0.15,
                                 end = 0.85)[c(1, 2, 4, 6, 7)]
```

```{r banner-data}
# data for banner plot
df <- data.frame(lineage = factor(lineage, levels = lineage), 
                 y = 1, 
                 functionality = functionality)
```

```{r banner, fig.height=.5, out.width = "100%", fig.align='center'}
# banner plot, acts as global key
df %>% 
  ggplot(aes(lineage, y, fill = lineage)) +
  geom_col(width = 1) +
  geom_label(aes(x = lineage, y = y*2/3, label = lineage),
             fill = "white",
             size = 2,
             label.size = 0,
             label.padding = unit(0.2, "lines")) +
  geom_label(aes(x = lineage, y = y*1/3, label = functionality),
             fill = "white",
             size = 2,
             label.size = 0,
             label.padding = unit(0.2, "lines")) +
  scale_fill_manual(values = line_colours) +
  theme_void() +
  theme(legend.position = "none")
```

# Introduction {#introduction}

Mesenchymal stromal cells (MSCs) are a subset non-hematopoietic adult stem cells that originate from the mesoderm. They have the capacity for self-renewal and differentiation into osteoblasts, adipocytes and chondrocytes making them important for the development of cell-based therapies in regenerative medicine. However, populations of MSCs are heterogeneous with respect to their differentiation capacity [@Costa2020-ze; @Phinney2012-yw] and some exhibit immundomodulatory properties unhelpful for cell-based therapies. Understanding subtype heterogeneity is key for the development of efficacious therapies. [The Genever Group](https://www.geneverlab.info/) has developed a model of this heterogeneity using has a number of immortalised clonal MSC lines which comprises five subtypes [@Stone2019-wr; @Kay2019-ec]. The phenotypes of these subtypes is as follows:

-   Y201 : the MSC stereotype which differentiate into osteoblasts, adipocytes and chondrocytes
-   Y101 : osteogenic capacity
-   Y101.5 : enhanced osteogenic capacity
-   Y202 : immunomodulatory capacity
-   Y102 : immunomodulatory capacity


The data are mass spectrometry data of the soluble protein fraction from five immortalised mesenchymal stromal cell (MSC) lines.

# Methods {#methods}

## Data description {#methods-data-description}

There are two data files: [Y101_Y102_Y201_Y202_Y101-5.csv](../data-raw/Y101_Y102_Y201_Y202_Y101-5.csv) and [comparison_p\_and_q.csv](../data-raw/comparison_p_and_q.csv)

The data in Y101_Y102_Y201_Y202_Y101-5.csv are normalised protein abundances. Each row is a protein and there are three replicates for each subtype which are arranged in columns. Also in the file are columns for:

-   the protein accessions: the top hit is first, Human sequences are prefixed with `1::`, bovine with `2::`
-   the number of peptides used to identify the protein
-   the number of unique peptides used to identify the protein
-   a measure of confidence in that identification
-   the maximum fold change between the mean abundances of two cell lines (i.e., highest mean / lowest mean)
-   a $p$ value for a one-way ANOVA for the effect of cell line
-   $q$ the false discovery rate, a $p$ value corrected by the Benjamini--Hochberg procedure [@Benjamini1995-yj].
-   a measure of the power of that test
-   the cell line with the highest mean
-   the cell line with the lowest mean
-   the protein mass
-   a description of the protein
-   a binary indicator variable for whether at least two peptides were detected for a protein.

The data in comparison_p\_and_q.csv give the p and q values for pairwise comparisons between cell line.

```{r data-import}
# file
filesol <- "../data-raw/Y101_Y102_Y201_Y202_Y101-5.csv"
# Column names are spread over three rows but are primarily in the third row.
# skip first two lines
sol <- read_csv(filesol, skip = 2) %>% 
  janitor::clean_names()

file <- "../data-raw/comparison_p_and_q.csv"
p_q <- read_csv(file) %>% 
  janitor::clean_names()
```

## Data preparation {#methods-data-preparation}

```{r data-filter}
# Remove irrelevant rows

# bovine serum proteins removed - these are from medium on which the cells were grown 
# proteins for which fewer than 2 peptides were detected
sol <- sol %>% 
  filter(str_detect(description, "OS=Homo sapiens")) %>% 
  filter(x1pep == "x")
num_prot <- length(sol$accession) 

p_q <- p_q %>% 
  filter(str_detect(description, "OS=Homo sapiens")) %>% 
  filter(x1pep == "x")

```

```{r add-useful-columns}
# Extract important information 'hidden' in other columns to columns of their own

# extract the genename from the description and put it in a column.
# genename is after GN= in the description until a white space
sol <- sol %>%
  mutate(genename =  str_extract(description,"GN=[^\\s]+") %>% 
           str_replace("GN=", ""))

p_q <- p_q %>%
  mutate(genename =  str_extract(description,"GN=[^\\s]+") %>% 
           str_replace("GN=", ""))


# Add a column, protid, for the top protein identifier. 
# This is the first Uniprot ID after the "1::" in the accession column.
sol <- sol %>%
  mutate(protid =  str_extract(accession, "1::[^;]+") %>% 
           str_replace("1::", ""))

```

```{r create-long-sol2}
# Create a second dataframe, sol2 in which the protein abundances are in a single column
# these are the columns beginning with a y
# the column names which indicate the cell lineage and replicate
# good to a column named lineage_rep which is then split into separate columns
# All the other variables are retained.
sol2 <- sol %>% pivot_longer(names_to = "lineage_rep",
                             values_to = "abundance",
                             cols = starts_with("y")) %>%
  extract(lineage_rep,
          c("line", "rep"),
          "(y[0-9]{3,4})\\_([a-c])")
```

```{r summary-sol}
# add a dataframe for the mean, se, sd, and n abundance for each cell line protein combination
sol_summary <- data_summary(sol2, 
                            measure = abundance, 
                            group1 = line,
                            group2 = protid)
# add a dataframe for the mean, se, sd, and n abundance for each cell line 
sol_summary_line <- data_summary(sol2, 
                            measure = abundance, 
                            group1 = line)
sol_summary_line$line <- factor(toupper(sol_summary_line$line), levels = lineage)
```

```{r augment-sol-summary}
# add the protein-line summary information to the main dataset
# requires protein-line summary info to be spread
sol_summary <- sol_summary %>% 
  pivot_wider(names_from = line,
              values_from = c(mean, n, sd, se))
# merge the dataframes
sol <- merge(sol, sol_summary, by = "protid")
```

```{r augment-sol-logfc}
# calculate the log fold changes for each of the comparisons
# we have p values for. This is all the pairwaise comparisions
# except that y1015 is commpared only to y101, no others
sol <- sol %>% 
  mutate(fc_y101_y1015 = log2(mean_y101/mean_y1015), # have p
         fc_y101_y102 = log2(mean_y101/mean_y102), # have p
         fc_y101_y201 = log2(mean_y101/mean_y201), # have p
         fc_y101_y202 = log2(mean_y101/mean_y202), # have p
         ##
         fc_y102_y201 = log2(mean_y102/mean_y201), # have p
         fc_y102_y202 = log2(mean_y102/mean_y202), # have p
         ##
         fc_y201_y202 = log2(mean_y201/mean_y202)) # have p
```


```{r protein-sets}
# a protein is present in a cell line if the mean abundance is greater than zero
protein_set <- list(Y1015 = sol$genename[sol$mean_y1015 > 0],
                    Y101 = sol$genename[sol$mean_y101 > 0],
                    Y201 = sol$genename[sol$mean_y201 > 0],
                    Y202 = sol$genename[sol$mean_y202 > 0],
                    Y102 = sol$genename[sol$mean_y102  > 0])

# number of proteins in each cell line
n_y1015 <- length(protein_set[["Y1015"]])
n_y101 <- length(protein_set[["Y101"]])
n_y102 <- length(protein_set[["Y102"]])
n_y201 <- length(protein_set[["Y201"]])
n_y202 <- length(protein_set[["Y201"]])
# this shows that y201 and y202 each have a protein missing from the 861

# the missing proteins are
absent_y201 <- setdiff(protein_set[["Y101"]],protein_set[["Y201"]])
absent_y202 <- setdiff(protein_set[["Y101"]],protein_set[["Y202"]])
```

I used R [@R-core] with **`tidyverse`** packages [@R-tidyverse] to import and process the raw data. Bovine serum proteins from the medium on which the cells were grown and those for which fewer than two peptides were detected were filtered out. Abundances were summarised for each line and for each line-protein combination and log fold change between cell lines for each protein were calculated. See Equation \@ref(eq:logfc). Protein sets were visualised with the **`VennDiagram`** package [@VennDiagram].


\begin{equation}
FC_{jj'} = log_{2}\left(\frac{mean_{j}}{mean_{j'}}\right)
(\#eq:logfc)
\end{equation}

where: 

* $mean_{j}$ is the mean protein abundance for a given protein in cell line $j$  
* $mean_{j'}$ is the mean protein abundance for a given protein in another cell line $j'$  
* such that $FC_{jj'}$ is the $log_{2}$ fold change between line $j$ and line $j'$ for a given gene.


## Visualisation with Principal Components Analysis {#methods-visualisation}

```{r pca-preparation}
# PCA requires the proteins to be in columns and the samples to be in rows
# Transpose the columns containing the abundance values for each sample
tsol <- sol %>% select(starts_with("y")) %>%
  t() %>% 
  data.frame()

# Use the genenames in `sol` to name the columns in `tsol`
# I'm using the genenames rather than the protein ids because they
# are more recognisable
names(tsol) <- sol$genename 

# The sample names of sol have become the row names of tsol
# put these in a column 'sample
tsol$sample <- row.names(tsol)

# process the sample name so we have the cell lineage in one column and the replicate in another
tsol <- tsol %>% 
  extract(sample, 
          c("lineage","rep"),
          "(y[0-9]{3,4})\\_([a-c])")
```

```{r pca-run}
pca <- tsol %>% 
  select(-lineage, -rep) %>%
  prcomp(scale. = TRUE)
```

Principal Components Analysis was used for dimensionality-reduction to allow visualisation the distance between samples. It was conducted on protein abundances scaled to unit variance to prevent highly abundant proteins dominating the analysis. The package **`GGally`** [@GGally] was used to produced pairwise scatter plots of the first six principal components.

## Clustering and heatmaps {#methods-clustering-and-heatmaps}

```{r cluster-filter}
row.names(sol) <- sol$genename

# select genes where at least one comparison differs significantly at the q < 0.001 level
sol_sigq_0.01 <- sol %>% 
  filter(q_value <= 0.01) %>% 
  select(starts_with("y")) 

# number of significant q < 0.01 values
n_sigq_0.01 <- nrow(sol_sigq_0.01)

# select genes where at least one comparison differs significantly at the q < 0.001 level
sol_sigq_0.05 <- sol %>% 
  filter(q_value <= 0.05) %>% 
  select(starts_with("y")) 

# number of significant q < 0.05 values but not less than 0.01
n_sigq_0.05 <- nrow(sol_sigq_0.05) 
n_sigq_0.05_0.01 <- n_sigq_0.05 - n_sigq_0.01
# number of significant p < 0.01 and <0.05
n_sigp_0.01 <- length(sol$anova_p[sol$anova_p < 0.01])
n_sigp_0.05 <- length(sol$anova_p[sol$anova_p < 0.05])
```

```{r cluster-prep}
# Matrix format
mat <- sol_sigq_0.01 %>% as.matrix()
```

Hierarchical clustering was conducted and heatmaps produced with the package **`heatmaply`** [@heatmaply] with interactivity provided by **`plotly`** [@plotly]. Hierarchical clustering used the complete linkage method and was performed for both proteins and samples on the proteins that differed significantly between at least two lines. Significance was determined at the $q < 0.01$ level where $q$ is the False Discovery Rate (FDR) adjusted $p$-value. FDR is a method of correcting for multiple comparisons by the Benjamini–Hochberg procedure [@Benjamini1995-yj]. 
The abundances were "$z$-score normalised" for each protein before clustering and heatmapping. $Z$-score normalisation is way to make comparisons between, and visualisations of, abundances on different scales. See Equation \@ref(eq:zscore).


\begin{equation}
z_{ijk} = \frac{x_{ijk} - \bar{x_{ij}}}{sd_{ij}}  
(\#eq:zscore)
\end{equation}

where:  

* $x_{ijk}$ is the abundance for protein $i$, cell line $j$ and replicate $k$  
* $\bar{x_{ij}}$ is the mean abundance for protein $i$ and cell line $j$ across replicates  
* $sd_{ij}$ is the standard deviation on the abundance for protein $i$ and cell line $j$ across replicates  
* such that $z_{ijk}$ is the $z$-score normalised abundance for protein $i$, cell line $j$ and replicate $k$

## Volcano plots {#methods-volcano-plots}

```{r volcano-prep}
sol <- merge(sol, p_q, by = "genename")
```

# Results {#results}

There were `r num_prot` human protein identified from more than two peptides All `r num_prot` were found in all five cells line with two exceptions: `r absent_y201` was not seen in Y201 and `r absent_y202` was not seen in Y202. These are likely to be missing by chance from all three replicates rather than biologically meaningful. See figure \@ref(fig:protein-venn-fig).
The total soluble protein content of each cell line was broadly similar. See figure \@ref(fig:line-abund-fig).


```{r protein-venn-write, include=FALSE}
# suppresses the writing of the log file
futile.logger::flog.threshold(futile.logger::ERROR,
                              name = "VennDiagramLogger")
# create venn diagram and write to file
venn.diagram(protein_set,
             filename = "figures/venn.png",
             imagetype = "png" ,
             height = 1000,
             width = 1000,
             units = "px",
             cat.col = line_colours,
             fill = line_colours,
             col = line_colours,
             cat.default.pos = "outer",
             lwd = 1,
             cex = 0.5,
             cat.cex = 0.6,
             cat.fontfamily = "sans",
             fontfamily = "sans",
             margin = 0.1,
             cat.dist = 0.25)
```

```{r protein-venn-fig, out.width="50%", fig.cap='(ref:protein-venn-fig)'}
file <- "figures/venn.png"
knitr::include_graphics(file)

```

<!-- figure legend -->

(ref:protein-venn-fig) Protein set shared by cells lines. Created with package **`VennDiagram`** [@VennDiagram]

<!-- end of fig legend -->



```{r line-abund-fig, fig.height=4, fig.width=4, fig.cap="(ref:line-abund-fig)"}
# ..
sol_summary_line %>% 
  ggplot() +
  geom_col(aes(x = line, y = mean, fill = line)) +
  geom_errorbar(aes(x = line,
                    ymin = mean - se,
                    ymax = mean + se),
                width = .5) +
  scale_x_discrete(labels = lineage,
                   name = "") +
  scale_fill_manual(values = line_colours) +
  scale_y_continuous(name = "Protein Abundance (units)",
                     expand = c(0, 0)) +
  theme_classic() +
  theme(legend.position = "none")

```

<!-- figure legend -->

(ref:line-abund-fig) Mean soluble protein content for each cell line. Error bars are $\bar{x} \pm 1 s.e.$

<!-- end of fig legend -->


## Principal Components analysis {#results-principal-components-analysis}

```{r pca-report}
# extract the eigenvalues
res <- summary(pca)[["importance"]]
# extract the % variance explained by first and first 6 PC
percent1 <- round(res["Cumulative Proportion","PC1"] * 100, 1)  
percent6 <- round(res["Cumulative Proportion","PC6"] * 100, 1)  
```

The first Principal Component captured `r percent1`% of the variation in soluble protein expression between samples and the first six captured `r percent6`%. The distributions of scores on PC1 for each cell lineage show good separation between Y1015 and the other cell lines. The two immunomodulatory lines, Y102 and Y202 are the least easy to separate on a single component. Pairwise scatter plots of PC1 to PC6 show greater, but still imperfect, separation of lines. See figure \@ref(fig:pca-vis-fig).

```{r pca-vis-prep1}
# the pairwise plots of PC with a legend are produced by adding an inset of the lgend to a figure of the pairwise plots. Since ggpairs ouputs a ggmatrix object rather than a ggplot object, this is save to file then placed as an image on a plot to which the inset is added

# extract the pca scores
pca_labelled <- data.frame(pca$x, 
                           lineage = factor(toupper(tsol$lineage),
                                            levels = lineage))
# add the functionality
pca_labelled <- merge(pca_labelled, df, by = "lineage")

# create the ggpairs fig (a ggmatrix object)
p1 <- pca_labelled %>% 
  select(lineage,
         PC1:PC6) %>%
  ggpairs(aes(color = lineage, fill = lineage),
          upper = NULL,
          columns = 2:7) + 
  scale_colour_manual(values = line_colours) +
  scale_fill_manual(values = line_colours) +
  theme_minimal()

# save to file
ggsave("figures/pca_pairs.png", 
       plot = p1, 
       device = "png",
       width = 5.8, 
       height = 5.5,
       units = "in",
       dpi = 300)
```

```{r pca-vis-2}
# import the pairs image
img <- png::readPNG("figures/pca_pairs.png")

# main plot comprising the pairs image
p1 <- df %>% ggplot(aes(x = lineage, y = y)) +
  annotation_custom(rasterGrob(img, 
                               width = unit(1,"npc"),
                               height = unit(1,"npc")), 
                    -Inf, Inf, -Inf, Inf) +
  theme_void()

# plot to act as inset legend
p2 <- df %>% 
  ggplot(aes(x = lineage, y = y, colour = lineage)) +
  geom_point(size = 8) +
  geom_text(aes(x = lineage, y = y*2/3, label = lineage),
            colour = "black") +
  scale_colour_manual(values = line_colours) +
  scale_y_continuous(limits = c(0, 1.1)) +
  theme_void() +
  coord_flip() +
  theme(legend.position = "none")

```

```{r pca-vis-fig, fig.width=8, fig.height=7.5,fig.cap="(ref:pca-vis-fig)"}
# final pca figure
p1 + inset_element(p2, 0.70, 0.68, 0.9, 0.9) 
```
<!-- figure legend -->

(ref:pca-vis-fig) Samples represented on the first six Principal Components capturing `r percent6`% of the variation in soluble protein expression between samples. The leading diagonal shows the distribution of scores on each component with pairwise scatter plots of Principal Components below the diagonal. Created with the **`GGally`** [@GGally] and **`patchwork`** [@patchwork] packages.

<!-- end of fig legend -->


## Clustering and heatmaps {#results-cluster-heat}

There were `r n_sigp_0.05` proteins with $p < 0.05$  and `r n_sigp_0.01` proteins with $p < 0.01$ but `r num_prot` tests implies `r num_prot * 0.05` and `r num_prot * 0.01` respectively are false positives. The FDR adjusted $p$-value (or $q$-value) of 0.01 implies that 1% of significant tests will result in false positives. The abundance of `r n_sigq_0.05` proteins differed significantly between at least two cell lines at the $q < 0.05$ level and between `r n_sigq_0.01` proteins at the $q < 0.01$ level. This implies only `r n_sigq_0.05 * 0.05` and `r n_sigq_0.01 * 0.01` false positives.

The distribution of $q$-values was right skewed (see Figure \@ref(fig:q-fig)) with a mean of `r round(mean(sol$q_value), 4)` and a median of `r round(median(sol$q_value), 4)`. This what we would expect to see.


```{r q-fig, fig.width=8, fig.height=4,fig.cap="(ref:q-fig)"}
# labelling for the figure giving the number of proteins with abundances significantly different between at least two lines at a particular significance level.
lab0.05 <- bquote(italic(n)~"="~.(n_sigq_0.05_0.01))
lab0.01 <- bquote(italic(n)~"="~.(n_sigq_0.01))

# distribution of q values
sol %>% ggplot(aes(q_value)) +
  geom_density() + 
  geom_vline(xintercept = 0.05) + 
  geom_vline(xintercept = 0.01) +
  scale_x_continuous(expand = c(0, 0),
                     name = "Density") +
  scale_y_continuous(expand = c(0, 0),
                     name =  expression(italic(q) * "-value (FDR corrected" ~ italic(p) * "-value)") ) +
  annotate("text", x = 0.15, y = 2, label = "Not significant") +
  annotate("text", x = 0.03, y = 2.4, label = expression("0.01<"~italic(p)~"< 0.05")) +
  annotate("text", x = 0.1, y = 6.4, label = expression(italic(p)~"< 0.01")) +
  annotate("text", x = 0.03, y = 2, label = lab0.05) +
  annotate("text", x = 0.1, y = 6, label = lab0.01) +
  geom_segment(x = 0.085, y = 6, xend = 0.005, yend = 5) +
  theme_classic()
```

<!-- figure legend -->

(ref:q-fig) Distribution of $q$-values ($p$-values corrected for multiple comparisons by the Benjamini–Hochberg procedure [@Benjamini1995-yj]).  

<!-- end of fig legend -->

```{r}
n_gen_clusters <- 3
n_cell_clusters <- 5
```

*to be completed*
Cell line replicates clustered by protein expression. These clustered corresponded 
The two immunomodulatory (see Figure \@ref(fig:heatmap-fig)) foorm one clus


```{r heatmap-fig, fig.width=8, fig.height=14,fig.cap="(ref:heatmap-fig)"}
heatmaply(mat, 
          scale = "row",
          grid_color = "white",
          hide_colorbar = TRUE,
          k_col = n_cell_clusters,
          k_row = n_gen_clusters,
          label_names = c("Protein", "Sample", "Normalised Abundance"),
          fontsize_row = 7, fontsize_col = 10,
          labCol = toupper(colnames(mat)),
          labRow = rownames(mat),
          heatmap_layers = theme(axis.line = element_blank()))
```

<!-- figure legend -->

(ref:heatmap-fig) my lovely heatmap

<!-- end of fig legend -->



## Volcano plots {#results-volcano-plots}

```{r}
sol %>% 
  ggplot(aes(x = fc_y101_y1015, 
             y = -log(q_y101_y1015),
             colour = abs(fc_y101_y1015) > 2 & q_y101_y1015 < 0.05)) +
  geom_point() +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
file <- "../sessioninfo.md"
writeLines(capture.output(sessionInfo()), file)
```

## Word count {#word-count}

Word count calculated with wordcountaddin [@wordcountaddin]. The session information was written into a separate file rather than to the README this has been added to the wordcount.

```{r wordcount}
wc_main <- wordcountaddin:::text_stats("main.Rmd")[3] %>%
  str_extract("[0-9]+") %>% 
  as.numeric()
wc_readme <- wordcountaddin:::text_stats("../README.md")[3] %>%
  str_extract("[0-9]+") %>%
  as.numeric()
wc_session <- wordcountaddin:::text_stats("../sessioninfo.md")[3] %>%
  str_extract("[0-9]+") %>%
  as.numeric()

```

This document: `r wc_main`\
README: `r wc_readme`\
Session info: `r wc_session`\
**Total: `r wc_main + wc_readme + wc_session`**

# References {#references}
